---
title: "ABDA: Assignment 5"
subtitle: Hierarchical Reaction Time Model
author: "Fredrik Nyström"
date: '2019-11-12'
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      css: ["hygge", "nfa.css"]
      ratio: "16:9"
      titleSlideClass: [left, bottom]
      highlightLines: true
      highlightStyle: solarized-light
      highlightSpans: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(rstan)
library(tictoc)
library(bayesboot)
library(bayestestR)
library(cowplot)
library(gridGraphics)
knitr::opts_chunk$set(echo = TRUE)
```
## Table of Contents

* [Task A.1](#a1)
* [Task A.2](#a2)

---
name: a1
## Data

```{r, echo = FALSE}
# Reaction times
y <- scan(file = here("data", "A5_y.csv"), sep = ",", quiet = TRUE)
# ID for the individuals
ind <- scan(file = here("data", "A5_ind.csv"), sep = ",", quiet = TRUE)
# My ID is 11

reaction_time_data <- list(
  N_data        = length(y),
  reaction_time = y,
  N_subjects    = length(unique(ind)),
  subjects      = ind
)
```

```{r}
glimpse(reaction_time_data)
```
---

## Model

```{stan output.var="a5", echo = FALSE, cache = TRUE}
data {
   int<lower = 0> N_data;
   vector[N_data] reaction_time;
   
   int<lower = 0> N_subjects;
   int subjects[N_data];
}

transformed data {
   vector[N_data] log_reaction_time = log(reaction_time);
}

parameters {
   // Group level
   real<lower = 0> mu;
   real<lower = 0> tau;
   // Individual level
   vector<lower = 0>[N_subjects] theta;
   real<lower = 0> sigma;
} 

model {
   theta ~ normal(mu, tau);
   for (i in 1:N_data) {
      log_reaction_time[i] ~ normal(theta[subjects[i]], sigma);
   }
}

generated quantities { 
   real<lower = 0> mu_exp = exp(mu + tau^2/2 + sigma^2/2);
   
   vector<lower = 0>[N_subjects] theta_exp;
   real<lower = 0> reaction_time_ppc;
   
   for (j in 1:N_subjects) {
      theta_exp[j] = exp(theta[j] + sigma^2/2);
   }
   
   reaction_time_ppc = exp(normal_rng(normal_rng(mu, tau), sigma));
}

```

```{stan output.var="a5a", eval = FALSE, echo = TRUE}
data {
   int<lower = 0> N_data;
   vector[N_data] reaction_time;
   
   int<lower = 0> N_subjects;
   int subjects[N_data];
}

transformed data {
   vector[N_data] log_reaction_time = log(reaction_time);
}

parameters {
   // Group level
   real<lower = 0> mu;
   real<lower = 0> tau;
   // Individual level
   vector<lower = 0>[N_subjects] theta;
   real<lower = 0> sigma;
} 

```

---

## Model cont.

```{stan output.var="a5b", eval = FALSE, echo = TRUE}
model {
   theta ~ normal(mu, tau);
   for (i in 1:N_data) {
      log_reaction_time[i] ~ normal(theta[subjects[i]], sigma);
   }
}

generated quantities { 
   real<lower = 0> mu_exp = exp(mu + tau^2/2 + sigma^2/2);
   
   vector<lower = 0>[N_subjects] theta_exp;
   real<lower = 0> reaction_time_ppc;
   
   for (j in 1:N_subjects) {
      theta_exp[j] = exp(theta[j] + sigma^2/2);
   }
   
   reaction_time_ppc = exp(normal_rng(normal_rng(mu, tau), sigma));
}

```

---
## Task A.1

What is the expected reaction time for the individual named "the dude" in our data set (index = 4)?

a) Answer this by providing a histogram and appropriate summaries.

```{r, echo = TRUE, results = "hide"}
stan_samples <- rstan::sampling(a5, data = reaction_time_data, iter = 8000)
```
```{r}
samples <- rstan::extract(stan_samples)

dude <- enframe(samples$theta_exp[, 4], name = NULL, value = "samples")
dude_hdi <- hdi(dude$samples, ci = 0.95)


dude_lbls <- c(
  Mean = paste0("Mean:   ", format(mean(dude$samples), digits = 3)),
  Median = paste0("Median: ", format(median(dude$samples), digits = 3)),
  MAP = paste0("MAP:    ", format(map_estimate(dude$samples), digits = 3)),
  HDI = paste0("HDI:    [", 
         format(dude_hdi$CI_low, digits = 3),", ",
         format(dude_hdi$CI_high, digits = 3), "]")
)
```

---
## Code for the histogram

```{r, eval = FALSE}
dude %>% ggplot(aes(x = samples)) + geom_histogram(binwidth = 5, alpha = 0.3) + 
  geom_vline(aes(xintercept = mean(dude$samples), color = "Mean")) + 
  geom_vline(aes(xintercept = map_estimate(dude$samples), color = "MAP")) + 
  geom_vline(aes(xintercept = median(dude$samples), color = "Median")) + 
  geom_segment(aes(x = dude_hdi$CI_low, xend = dude_hdi$CI_high, y = 0, yend = 0, color = "HDI"), size = 2) + 
  scale_color_manual(name = "Estimates", labels = dude_lbls, 
                     values = c(Mean = "blue", MAP = "green", Median = "red", HDI = "black")) +
  theme_bw() + theme(legend.position = c(0.9, 0.85)) + 
  labs(title = "Reaction time samples from the Dude",
       subtitle = paste(length(dude$samples), "samples."),
       x = "Reaction time (ms)", y = "Number of samples")
```
---
## The histogram

```{r, echo = FALSE}
dude %>% ggplot(aes(x = samples)) + geom_histogram(binwidth = 5, alpha = 0.3) + 
  geom_vline(aes(xintercept = mean(dude$samples), color = "Mean")) + 
  geom_vline(aes(xintercept = map_estimate(dude$samples), color = "MAP")) + 
  geom_vline(aes(xintercept = median(dude$samples), color = "Median")) + 
  geom_segment(aes(x = dude_hdi$CI_low, xend = dude_hdi$CI_high, y = 0, yend = 0, color = "HDI"), size = 2) + 
  scale_color_manual(name = "Estimates", labels = dude_lbls, 
                     values = c(Mean = "blue", MAP = "green", Median = "red", HDI = "black")) +
  theme_bw() + theme(legend.position = c(0.85, 0.85)) + 
  labs(title = "Reaction time samples from the Dude",
       subtitle = paste(length(dude$samples), "samples."),
       x = "Reaction time (ms)", y = "Number of samples")
```

---
## Task A.1.b

With only one measurement, how are we able to give a 95% credible interval for "the dude"? This is not possible in frequentist statistics.

In the hierarchical model all individuals share the same $\sigma$, so it is possible to get samples from the dude since the model utilizes the information from all other invidividuals in calculating the $\sigma$.

---
name: a2
## Task A.2

What is the group's reaction time?

a) Given a new individual from our group, what is the expected
   i) reaction time?
   
```{r}
mean(samples$mu_exp)
hdi(samples$mu_exp, ci = 0.95)
```

---
## Task A.2
   ii) predicted reaction time for a single measurement for that individual?
```{stan output.var="a5c", eval = FALSE, echo = TRUE}
generated quantities { 
   real<lower = 0> mu_exp = exp(mu + tau^2/2 + sigma^2/2);
   
   vector<lower = 0>[N_subjects] theta_exp;
   real<lower = 0> reaction_time_ppc; #<<
   
   for (j in 1:N_subjects) {
      theta_exp[j] = exp(theta[j] + sigma^2/2);
   }
   
   reaction_time_ppc = exp(normal_rng(normal_rng(mu, tau), sigma)); #<<
}

```
---
## Task A.2 Estimates
   ii) predicted reaction time for a single measurement for that individual?
 
```{r}
random <- enframe(samples$reaction_time_ppc, name = NULL, value = "samples")
random_hdi <- hdi(random$samples, ci = 0.95)

rnd_lbls <- c(
  Mean = paste0("Mean:   ", format(mean(random$samples), digits = 3)),
  Median = paste0("Median: ", format(median(random$samples), digits = 3)),
  MAP = paste0("MAP:    ", format(map_estimate(random$samples), digits = 3)),
  HDI = paste0("HDI:    [", 
               format(random_hdi$CI_low, digits = 3),", ",
               format(random_hdi$CI_high, digits = 3), "]")
)
```
---
## Task A.2. Histogram code

```{r, echo = TRUE, eval = FALSE}
random %>% ggplot(aes(x = samples)) + geom_histogram(binwidth = 5, alpha = 0.3) + 
  geom_vline(aes(xintercept = mean(random$samples), color = "Mean")) + 
  geom_vline(aes(xintercept = map_estimate(random$samples), color = "MAP")) + 
  geom_vline(aes(xintercept = median(random$samples), color = "Median")) + 
  geom_segment(aes(x = random_hdi$CI_low, xend = random_hdi$CI_high, y = 0, yend = 0, color = "HDI"), size = 2) + 
  scale_color_manual(name = "Estimates", labels = rnd_lbls, 
                     values = c(Mean = "blue", MAP = "green", Median = "red", HDI = "black")) +
  theme_bw() + theme(legend.position = c(0.9, 0.85)) + 
  labs(title = "Reaction time samples from new random individual",
       subtitle = paste(length(random$samples), "samples."),
       x = "Reaction time (ms)", y = "Number of samples")
```
---
## Task A.2. Histogram
.pull-left[
```{r, echo = FALSE}
random %>% ggplot(aes(x = samples)) + geom_histogram(binwidth = 5, alpha = 0.3) + 
  geom_vline(aes(xintercept = mean(random$samples), color = "Mean")) + 
  geom_vline(aes(xintercept = map_estimate(random$samples), color = "MAP")) + 
  geom_vline(aes(xintercept = median(random$samples), color = "Median")) + 
  geom_segment(aes(x = random_hdi$CI_low, xend = random_hdi$CI_high, y = 0, yend = 0, color = "HDI"), size = 2) + 
  scale_color_manual(name = "Estimates", labels = rnd_lbls, 
                     values = c(Mean = "blue", MAP = "green", Median = "red", HDI = "black")) +
  theme_bw() + theme(legend.position = c(0.9, 0.85)) + 
  labs(title = "Reaction time samples from new random individual",
       subtitle = paste(length(random$samples), "samples."),
       x = "Reaction time (ms)", y = "Number of samples")
```
]

.pull-right[
iii) Compare the PPC histogram with Website data: 

median = 273 ms

mean = 284 ms
]
---
## Task A.3
Compare the thetas from the Bayesian hierarchical model with the sample means from the individuals.

```{r, results = "hide", warnings = FALSE}
group_stats <- tibble(ind = 1:34, group_mean = mean(samples$mu_exp))

sample_stats <- tibble(time = y, ind = ind) %>% 
  group_by(ind) %>% 
  summarize(mean = mean(time), sd = sd(time))

thetas <- as_tibble(samples$theta_exp) %>% 
  pivot_longer(everything(), 
               names_to = "ind", values_to = "time", 
               names_pattern = "V(\\d+)") %>%
  mutate(ind = ordered(ind, levels = 1:34))
```

---
## Task A.3 Histogram code

```{r, eval = FALSE}
thetas %>% ggplot(aes(x = time)) + geom_histogram(binwidth = 5, alpha = 0.3) + 
   geom_vline(data = sample_stats, aes(xintercept = mean)) + 
   geom_vline(data = group_stats, aes(xintercept = group_mean), color = "blue") +
   facet_wrap(~ ind) +
   theme_bw()
```
---
## Task A.3. Histogram
Black histogram = sampled reaction time
Black = calculated sample mean
Blue = Sampled group mean

```{r, echo = FALSE, fig.width=15}
thetas %>% ggplot(aes(x = time)) + 
   geom_histogram(binwidth = 5, alpha = 0.3) + 
   geom_vline(data = sample_stats, aes(xintercept = mean)) + 
   geom_vline(data = group_stats, aes(xintercept = group_mean), color = "blue") +
   facet_wrap(~ ind, ncol = 7) +
   theme_bw()
```

